---
title: 'Health inequity seen through the lens of infant mortality'
author: "SDS322E"
date: ''
output:
  html_document:
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
  pdf_document:
    toc: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, fig.align = "center", warning = F, message = F,
tidy=TRUE, tidy.opts=list(width.cutoff=60), R.options=list(max.print=100))
```
#### Austin Ivery-Clemons

## Introduction
My project contains the following datasets: "zipcode_imr", "zipcode_dem", and "zipcode_income". The "zipcode_imr" dataset is from *The University of Texas System* and was originally titled "*Infant Mortality in Communities Across Texas*". The dataset contains every zip code in the state of Texas along with the infant mortality rate (death per 1,000 births) for that zip code, if applicable. The two variables for this dataset are zip codes and infant mortality rates. The "zipcode_dem" dataset is from *data.census.gov* and is a demogrpahic table created by the US census. The dataset contains demographic data from every zip code in Texas, and I was able to download specific zip codes for my data. The variables for this dataset included total population, median age, number of hispanic individuals of any race, number of white individuals, number of black individuals, number of Native American individuals, and number of Asian individuals for each selected zip code. The final dataset was the "zipcode_income" dataset. The dataset is also from *data.census.gov* from the US census. This dataset contains the median household income for every zip code in Texas, and I was able to download specific zip codes for my data. The variable for this dataset included the median household income for each dataset specified.

The reason chose these dataset is due to my interest in health equity. I am a biology and black studies major and plan to attend medical school. I am also a part of a research team currently working towards health equity. I wanted to do a project revolved around my passion, which is why I chose infant mortality rates around Texas. This project will put health inequity into numbers. It will show that areas with less money, and with more black and brown individuals, endure more infant mortality than areas with more money. To visualize helath inequity is important, because a problem must first be recognized in order for it to be fixed, and for me, this project aids in recognizing the problem. The zip codes analyzed for this project were predominantly from Austin and Houston. All of Austin's zip codes are included, and some zip codes around Austin as well. The zip codes from Houston have been specifically chosen to include predominantly black areas. This was to create a difference in demographics, and be able to compare between predominantly white areas to predominantly black and brown areas.

---

```{R}
library(tidyverse)

zipcode_imr <- read_csv("zipcode_imr.csv")
zipcode_imr %>% glimpse()

zipcode_dem <- read_csv("zipcode_dem.csv", 
    col_names = FALSE)
zipcode_dem %>% glimpse()

zipcode_income <- read_csv("zipcode_income.csv", 
    col_names = FALSE)
zipcode_income %>% glimpse()
```

---

## Tidying: Reshaping

##### Reshaping dataset: "zipcode_imr":
```{r}
zipcode_imr <- zipcode_imr %>% pivot_wider(names_from = "Zip", values_from = "IMR")
zipcode_imr <- zipcode_imr %>% pivot_longer(1:2455, names_to = "zip_code", values_to = "IMR")
glimpse(zipcode_imr)
```

##### Reshaping dataset: "zipcode_dem":
```{R}
zipcode_dem <- zipcode_dem %>% pivot_longer(cols = c('X2':'X265'))
zipcode_dem <- zipcode_dem %>% pivot_wider(names_from = "X1", values_from = "value")
zipcode_dem <- zipcode_dem %>% select(-name)
zipcode_dem <- zipcode_dem %>% separate(Label, into = c(NA, "zip_code"), sep = 6)
zipcode_dem <- zipcode_dem %>% filter(!str_detect(zip_code, "Margin"), !str_detect(zip_code, "Percent"))
zipcode_dem <- zipcode_dem %>% separate(zip_code, into = c("zip_code", NA), sep = 5)

zipcode_dem <- zipcode_dem %>% select(zip_code, total_population = "Total population", median_age = "Median age (years)", hispanic_latino_any_race = "Hispanic or Latino (of any race)", white_alone = "White alone", black_alone = "Black or African American alone", native_american_alone = "American Indian and Alaska Native alone", asian_alone = "Asian alone")

zipcode_dem <- as.data.frame(apply(zipcode_dem, 2, as.numeric))
zipcode_dem$zip_code <- as.character(zipcode_dem$zip_code)

glimpse(zipcode_dem)
```

##### Reshaping dataset: "zipcode_income":
```{r}
zipcode_income <- zipcode_income %>% pivot_longer(cols = c('X2':'X537'))
zipcode_income <- zipcode_income %>% pivot_wider(names_from = "X1", values_from = "value")
zipcode_income <- zipcode_income %>% select(-name, -Total, -"Mean income (dollars)")

zipcode_income <- zipcode_income %>% filter(!str_detect(Label, "Texas"), !str_detect(Label, "Margin"), !str_detect(Label, "Families"), !str_detect(Label, "families"), !str_detect(Label, "Nonfamily"))

zipcode_income <- zipcode_income %>% separate(Label, into = c(NA, "zip_code"), sep = 6)
zipcode_income <- zipcode_income %>% separate(zip_code, into = c("zip_code", NA), sep = 5)
zipcode_income <- zipcode_income %>% select(zip_code, median_income = "Median income (dollars)")
zipcode_income$median_income <- as.numeric(zipcode_income$median_income)

glimpse(zipcode_income)
```
The dataset "zipcode_imr" was already tidy so here it has been untidied with 'pivot_wider()' and then tidied again with 'pivot_longer()'. The dataset 'zipcode_dem' was not tidy, so it was tidied with 'pivot_long()' and then cleaned up with various functions. The dataset 'zipcode_income' was also not tidy. It was tidied with 'pivot_longer()' and then also cleaned up with various functions. 

---

## Joining/Merging

##### Find what ID's in "zipcode_dem" that don't appear in "zipcode_imr":
```{r}
anti_join(zipcode_dem, zipcode_imr, by = "zip_code")
```

##### Find what ID's in "zipcode_income" that don't appear in "zipcode_imr":
```{r}
anti_join(zipcode_income, zipcode_imr, by = "zip_code")
```

##### Find what ID's in "zipcode_imr" don't appear in "zipcode_dem":
```{r}
anti_join(zipcode_imr, zipcode_dem, by = "zip_code")
```

##### Find what ID's in "zipcode_imr" don't appear in "zipcode_income":
```{r}
anti_join(zipcode_imr, zipcode_income, by = "zip_code")
```

##### Find what ID's don't appear between "zipcode_dem" and "zipcode_income":
```{r}
anti_join(zipcode_dem, zipcode_income, by = "zip_code")
anti_join(zipcode_income, zipcode_dem, by = "zip_code")
```

##### Joining:
```{R}
zipcode_joined <- zipcode_dem %>% inner_join(zipcode_imr, by = "zip_code")
zipcode_joined <- zipcode_joined %>% inner_join(zipcode_income, by = "zip_code")

glimpse(zipcode_joined)
```

An "inner_join()" will be used. The "zipcode_dem" and "zipcode_income" datasets both have 66 observations/rows each, and the "zipcode_imr" dataset has 2,455 observations/rows and the "zipcode_income" dataset has 66 observations/rows. Each ID is unique in all three datasets, so "zipcode_dem" and "zipcode_income" have 66 unique IDs and "zipcode_imr" has 2,455 unique IDs. All IDs in the "zipcode_dem" and "zipcode_income" datasets appeared in the "zipcode_imr" dataset. 2,389 IDs, or zip codes, appeared in the "zipcode_imr" dataset that didn't appear in the "zipcode_dem" or "zipcode_income" datasets. The ID's between "zipcode_dem" and "zipcode_imr" are exactly the same, there is no difference between their ID's. 0 observations were dropped from "zipcode_dem" and "zipcode_income", while 2,389 observations were dropped from "zipcode_imr". The reason I chose an inner_join is because the "zipcode_dem" dataset, which contains the demographic data, and the "zipcode_income" dataset, which contains the median income data, both only had 66 observations. Since the "zipcode_imr" dataset contains every Texas zip code, but not demographic nor income data, the only data I could use to analyzed would be the zip codes that only appear in all three, and not the extra zip codes in the "zipcodes_imr" dataset that don't appear in "zipcodes_dem" and "zipcode_income" datasets. The joined dataset now has 66 observations/rows, which is the exact same as the "zipcode_dem" and "zipcode_income" datasets, but considerably less than the "zipcode_imr" dataset. A potential problem with dropping many zip codes from the "zipcode_imr" dataset is that only certain areas of Texas are being analyzed rather than the entirety of Texas, so the data could skew and misrepresent the rest of Texas zip codes not being represented here. 

---



## Wrangling

"*filter()*" and "*str_detect()*" were both utilized in the tidying section to in order to manipulate the data and discard of rows containing unnecessary data with the following code (seen above):

_*zipcode_dem %>% filter(!str_detect(zip_code, "Margin"), !str_detect(zip_code, "Percent"))*_

_*zipcode_income %>% filter(!str_detect(Label, "Texas"), !str_detect(Label, "Margin"), !str_detect(Label, "Families"), !str_detect(Label, "families"), !str_detect(Label, "Nonfamily"))*_




##### Utilizing mutate to create categorical variable:
```{r}
zipcode_joined <- zipcode_joined %>% mutate(income_class = case_when(median_income < 35000 ~ "lower class",
                                                   median_income >= 35000 & 50000 > median_income ~ "lower middle class",
                                                   median_income >= 50000 & 110000 >= median_income ~ "middle class",
                                                   median_income > 110000 ~ "upper middle class"))

zipcode_joined %>% glimpse()
```

##### Utilizing mutate to create new variable that is function of another:
```{r}
zipcode_joined <- zipcode_joined %>% mutate(proportion_hispanic = hispanic_latino_any_race/total_population,
                          proportion_white = white_alone/total_population,
                          proportion_black = black_alone/total_population,
                          proportion_native = native_american_alone/total_population,
                          proportion_asian = asian_alone/total_population)

zipcode_joined %>% glimpse
```

##### Utilizing select in order to arrange columns differently:
```{r}
zipcode_joined <- zipcode_joined %>% select(zip_code, total_population, median_age, median_income, income_class, IMR, hispanic_latino_any_race, proportion_hispanic, white_alone, proportion_white, black_alone, proportion_black, native_american_alone, proportion_native, asian_alone, proportion_asian)

zipcode_joined %>% glimpse()
```

##### Utilizing arrange to find zip codes arranged by IMR, from greatest IMR to least IMR:
```{r}
zipcode_joined %>% arrange(desc(IMR))
```


#### Summary statistics:

##### Income stats
```{r}
zipcode_joined %>% summarize(mean_inc = mean(median_income, na.rm = T), sd_inc = sd(median_income), var_inc = var(median_income), min_inc = min(median_income), max_inc = max(median_income), distinct_inc = n_distinct(median_income), median_ = median(median_income)) %>% knitr::kable()

zipcode_joined %>% group_by(income_class) %>% summarize(mean_inc = mean(median_income, na.rm = T), sd_inc = sd(median_income), var_inc = var(median_income), min_inc = min(median_income), max_inc = max(median_income), distinct_inc = n_distinct(median_income), median_ = median(median_income)) %>% knitr::kable()
```

##### IMR stats
```{r}
zipcode_joined %>% summarize(mean_IMR = mean(IMR, na.rm = T), sd_IMR = sd(IMR, na.rm = T), var_IMR = var(IMR, na.rm = T), min_IMR = min(IMR, na.rm = T), max_IMR = max(IMR, na.rm = T), distinct_IMR = n_distinct(IMR), median_IMR = median(IMR, na.rm = T)) %>% knitr::kable()

zipcode_joined %>% group_by(income_class) %>% summarize(mean_IMR = mean(IMR, na.rm = T), sd_IMR = sd(IMR, na.rm = T), var_IMR = var(IMR, na.rm = T), min_IMR = min(IMR, na.rm = T), max_IMR = max(IMR, na.rm = T), distinct_IMR = n_distinct(IMR), median_IMR = median(IMR, na.rm = T)) %>% knitr::kable()
```

##### Age statistics
```{r}
zipcode_joined %>% summarize(mean_age = mean(median_age, na.rm = T), sd_age = sd(median_age, na.rm = T), var_age = var(median_age, na.rm = T), min_age = min(median_age, na.rm = T), max_age = max(median_age, na.rm = T), distinct_age = n_distinct(median_age), median_age = median(median_age, na.rm = T)) %>% knitr::kable()

zipcode_joined %>% group_by(income_class) %>% summarize(mean_age = mean(median_age, na.rm = T), sd_age = sd(median_age, na.rm = T), var_age = var(median_age, na.rm = T), min_age = min(median_age, na.rm = T), max_age = max(median_age, na.rm = T), distinct_age = n_distinct(median_age), median_age = median(median_age, na.rm = T)) %>% knitr::kable()
```

##### Hispanic individuals statistics
```{r}
zipcode_joined %>% summarize(mean_hisp = mean(proportion_hispanic, na.rm = T), sd_hisp = sd(proportion_hispanic), var_hisp = var(proportion_hispanic), min_hisp = min(proportion_hispanic), max_hisp = max(proportion_hispanic), distinct_hisp = n_distinct(proportion_hispanic), median_hisp = median(proportion_hispanic)) %>% knitr::kable()

zipcode_joined %>% group_by(income_class) %>% summarize(mean_hisp = mean(proportion_hispanic, na.rm = T), sd_hisp = sd(proportion_hispanic), var_hisp = var(proportion_hispanic), min_hisp = min(proportion_hispanic), max_hisp = max(proportion_hispanic), distinct_hisp = n_distinct(proportion_hispanic), median_hisp = median(proportion_hispanic)) %>% knitr::kable()
```

##### White individuals statistics
```{r}
zipcode_joined %>% summarize(mean_white = mean(proportion_white, na.rm = T), sd_white = sd(proportion_white), var_white = var(proportion_white), min_white = min(proportion_white), max_white = max(proportion_white), distinct_white = n_distinct(proportion_white), median_white = median(proportion_white)) %>% knitr::kable()

zipcode_joined %>% group_by(income_class) %>% summarize(mean_white = mean(proportion_white, na.rm = T), sd_white = sd(proportion_white), var_white = var(proportion_white), min_white = min(proportion_white), max_white = max(proportion_white), distinct_white = n_distinct(proportion_white), median_white = median(proportion_white)) %>% knitr::kable()
```

##### Black individuals statistics
```{r}
zipcode_joined %>% summarize(mean_black = mean(proportion_black, na.rm = T), sd_black = sd(proportion_black), var_black = var(proportion_black), min_black = min(proportion_black), max_black = max(proportion_black), distinct_black = n_distinct(proportion_black), median_black = median(proportion_black)) %>% knitr::kable()

zipcode_joined %>% group_by(income_class) %>% summarize(mean_black = mean(proportion_black, na.rm = T), sd_black = sd(proportion_black), var_black = var(proportion_black), min_black = min(proportion_black), max_black = max(proportion_black), distinct_black = n_distinct(proportion_black), median_black = median(proportion_black)) %>% knitr::kable()
```

##### Native American individuals statistics
```{r}
zipcode_joined %>% summarize(mean_native = mean(proportion_native, na.rm = T), sd_native = sd(proportion_native), var_native = var(proportion_native), min_native = min(proportion_native), max_native = max(proportion_native), distinct_native = n_distinct(proportion_native), median_native = median(proportion_native)) %>% knitr::kable()

zipcode_joined %>% group_by(income_class) %>% summarize(mean_native = mean(proportion_native, na.rm = T), sd_native = sd(proportion_native), var_native = var(proportion_native), min_native = min(proportion_native), max_native = max(proportion_native), distinct_native = n_distinct(proportion_native), median_native = median(proportion_native)) %>% knitr::kable()
```

##### Asian individuals statistics
```{r}
zipcode_joined %>% summarize(mean_asian = mean(proportion_asian, na.rm = T), sd_asian = sd(proportion_asian), var_asian = var(proportion_asian), min_asian = min(proportion_asian), max_asian = max(proportion_asian), distinct_asian = n_distinct(proportion_asian), median_asian = median(proportion_asian)) %>% knitr::kable()

zipcode_joined %>% group_by(income_class) %>% summarize(mean_asian = mean(proportion_asian, na.rm = T), sd_asian = sd(proportion_asian), var_asian = var(proportion_asian), min_asian = min(proportion_asian), max_asian = max(proportion_asian), distinct_asian = n_distinct(proportion_asian), median_asian = median(proportion_asian)) %>% knitr::kable()
```

##### Function created to find standard error of average IMR, grouped by income class:
```{r}
zipcode_joined %>% group_by(income_class) %>% summarize(mean_imr = mean(IMR, na.rm = T),
                                                                      sd_imr = sd(IMR, na.rm = T), n =n(),
                                                                      se_imr = sd_imr/sqrt(n)) %>%
   knitr::kable()
```

##### Table of counts for income class (categorical variable):
```{r}
zipcode_joined %>% group_by(income_class) %>% summarize(count = n()) %>% knitr::kable()
```

##### The only variable with NAs is the 'IMR' variable. There are 8 NAs in the IMR variable.
```{r}
zipcode_joined %>% filter(is.na(IMR)) %>% knitr::kable()
```
'Filter()' and 'str_detect()' were utilized in conjunction to get ride of rows that had unecessary information. This was done during the tidying process and by manipulating the data it helped clean up the dataset. 'Mutate()' was utilized in order to create a categorical variable which groups median incomes into "lower class", "lower middle class", "middle class", or "upper middle class". 'Mutate()' was also used in order to create the proportion of the race deomogrpahics. I did this in order to compare demographics easier, and felt it was important to know the proportion of a population a race makes up rather than just the number itself. 'Select()' was then used to rearrange the columns, and 'arrange()' was used to order the zip codes from greatest to least in infant mortality rate. The statistics of each numeric was then found, and also was grouped by income class. A function was then created to find the standard error of average IMR, when grouped by income class. 13 zip codes were grouped in lower class due to their median income, 17 for lower middle class, 29 in middle class, and 7 in upper middle class. Finally, there were 8 NAs iin the IMR variable, which was the only variable which contained any NAs. 

It was interesting to see how big of a difference there was in IMR between income classes. Lower class zip codes had an average IMR of 6.5 while upper middle class zip codes had an average IMR of 2.5. That is a huge difference in IMR, with 4 more infants dying per 1,000 live births in poorer areas. It was also interesting to see the distribution of race between income classes. Upper middle class zip codes had an average proportion of white people of 0.80, while the lower class had an average proportion of 0.19. This a stark contrast to black and brown areas. Upper middle class zip codes had an average proportion of black people of 0.01 while the lower class had an average proportion of 0.32. For hispanics, the upper middle class average was 0.09 while the lower class average was 0.46. There is a considerable difference in the distribution of demographics when controlling for median household income. The trend shows the black and brown people live in poorer areas than white people, and pooere areas have a larrger infant mortality rate than wealthier areas.

---

## Visualizing

```{R}
zipcode_joined %>% na.omit() %>% group_by(income_class) %>% summarize(mean_imr = mean(IMR, na.rm = T),
                                                                      sd_imr = sd(IMR, na.rm = T), n =n(),
                                                                      se_imr = sd_imr/sqrt(n)) %>%
  ggplot(aes(x = income_class, y = mean_imr))+
  geom_bar(stat="summary", fun=mean) +
  geom_errorbar(aes(y = mean_imr, ymin = mean_imr-se_imr, ymax = mean_imr+se_imr)) +
  ggtitle("IMR of each income class") + xlab("Income class") + ylab("Average infant mortality rate") +
  theme(axis.text.x=element_text(angle=30, hjust=1)) + 
  scale_y_continuous(breaks=seq(0,10,1))
```

The plot depicts the average infant mortality rate seen in each income class. It is apparent that the lower class suffers from more infant deaths than the following, welathier areas. This graph depicts that trend that wealthier neighborhoods have lower infant deaths than areas relatively poorer. It is reinforced by each bar, which has a lower IMR than the one preceding it. 

```{R}
zipcode_joined %>% na.omit() %>%
  ggplot(aes(x = proportion_black, y = IMR)) +
  geom_point(aes(color = income_class)) +
  geom_smooth(method="lm") + 
  ggtitle("IMR vs. black population proportion") + xlab("Black population proportion") + ylab("Infant mortality rate") +
  theme(legend.position=c(.8,.2)) +
  scale_x_continuous(breaks=seq(0,1,.1))

zipcode_joined %>% na.omit() %>%
  ggplot(aes(x = proportion_white, y = IMR)) +
  geom_point(aes(color = income_class)) +
  geom_smooth(method="lm") +
  ggtitle("IMR vs. white population proportion") + xlab("White population proportion") + ylab("Infant mortality rate") +
  theme(legend.position=c(.8,.8)) +
  scale_x_continuous(breaks=seq(0,1,.1))
```

These two plots are the exact same kind of plots, however they depict a strong message between race and infant mortality. The graphs show that areas with increasing proportions of black residents is positively correlated with an increase in infant mortality rates. This is contrasted by the second graph, which depicts that areas with increasing proportion of white residents is negatively correlated with infant mortality rates. As the number of white residents increase, the rate of infant deaths decreases, which is the opposite of what is seen with black residents. 

```{R}
zipcode_joined %>% 
  ggplot(aes(x = IMR, fill = income_class)) +
  geom_density(alpha = .75) +
  geom_rug() +
  ggtitle("IMR density of each income class") + xlab("Infant mortality rate") + ylab("Density") +
  theme(legend.position=c(.9,.8)) +
  scale_x_continuous(breaks=seq(0,15,1))
```

This plot depicts the IMR density in each income class. It shows that the upper middle class has a high density of low IMR, the middle class has a high density of middle IMR, the lower middle class has a high density of a higher IMR, and the lower class has a spreaded out density of various IMRs and is the only income class with any desnity past an IMR of around 9. This plot reinforces that the wealthier an area is, the lower the infant mortality rate will be.